---
title: 'Wikimedia Commons: File Contributions by Bots and Users (Redux)'
author: "Megan Neisler"
date: "1/30/2018"
output:
  html_document: default
  'html_document: default': default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(0)
library(tidyverse)
```


```{r}
message("Create an auto-closing SSH tunnel in the backgroud")
system("ssh -f -o ExitOnForwardFailure=yes stat1006.eqiad.wmnet -L 3307:analytics-store.eqiad.wmnet:3306 sleep 10")
library(RMySQL)
con <- dbConnect(MySQL(), host = "127.0.0.1", group = "client", dbname = "commonswiki", port = 3307)
```

Epic Ticket: [T185363](https://phabricator.wikimedia.org/T185363)

# Baseline Metrics for SDoC: Reviewing File Contributions by Bots and Users 

This is a replicate of Chelsy's report from October 12, 2017, comparing file contributions by bots and users to Wikimedia Commons from lanunch of Commons, September 2004, to the end of January 2018.

On February 1, 2018, I queried the image table of the Wikimedia Commons database to find the number of files uploaded by users and bots (tools or operated by institutions. The query identified user accounts identfied as being in the bot group in user_groups table or user_former_groups table, and user accounts that belong to categories whose name match _(bot_flag|bots)(_|$), as bot accounts. I then broke these file counts down by media type (defined by img_media_type field in image table).  In addition, I reviewed the total number of file contributions each month including the cumulative sum of file uploads and files newly uploaded each month (non-cumulative).

The total number of files uploaded by bots is 9,418,169 (22.04%), and the number of files uploaded by users is 35,180,917 (77.96%). In just the month of January 2018, 10,066 new files were uploaded by bots and 653,997 new files were uploaded by users. This is an increase from the month of December 2017 where 7,831 new files were uploaded by bots and 498,542 new files were uploaded by users. 

Bitmaps are the most common media type files uploaded by both bots and users, comprising 94.30% of all media contributions to Wikimedia Commons.   


Ticket: [T185365](https://phabricator.wikimedia.org/T185365)


## File contributions by media type 

```{r}
query <- "
SELECT img_media_type, user_group, COUNT(*) AS n_files
FROM (
-- Get active bots
SELECT ug_user AS user_id, ug_group AS user_group
FROM user_groups
WHERE ug_group = 'bot'
-- Get inactive bots
UNION
SELECT ufg_user AS user_id, ufg_group AS user_group
FROM user_former_groups
WHERE ufg_group = 'bot'
UNION
-- Get user ids with bot categories in their user pages
SELECT user.user_id, 'bot' AS user_group
FROM user INNER JOIN (
  -- select all user page names with bot category
  SELECT REPLACE(page.page_title, '_', ' ') AS user_name
  FROM page INNER JOIN (
    -- select page ids with bot categories 
    SELECT DISTINCT cl_from AS page_id
    FROM categorylinks
    WHERE cl_to REGEXP '_(bot_flag|bots)(_|$)'
      AND cl_type = 'page'
  ) AS bot_cat ON page.page_id=bot_cat.page_id
  WHERE page_namespace = 2
) AS bot_name ON user.user_name=bot_name.user_name
) AS bots RIGHT JOIN image ON bots.user_id = image.img_user
GROUP BY img_media_type, user_group;
"
```


```{r}

message("Using SSH tunnel & connection to Analytics-Store...")
bot_by_mediatype <- wmf::mysql_read(query, "commonswiki", con = con)
message("Closing connection...")
wmf::mysql_close(con)
save(bot_by_mediatype, file = "data/bot_by_mediatype.RData")
```

### File contributions by all media types
```{r}
load("data/bot_by_mediatype.Rdata")

bot_by_mediatype <- bot_by_mediatype %>%
        mutate(user_group = if_else(is.na(user_group), "user", user_group),
               img_media_type = tolower(img_media_type),
               proportion = paste0(round(n_files/sum(n_files), 4)*100, "%"))


bot_by_mediatype %>%
        group_by(user_group) %>%
        summarize(n_files = sum(n_files)) %>%
        mutate(proportion = n_files/sum(n_files))


knitr::kable(bot_by_mediatype)
```

### File contributions by non-bitmap media types
```{r}
bot_by_mediatype_omitbitmap <- bot_by_mediatype[bot_by_mediatype$img_media_type != "bitmap",]

bot_by_mediatype_omitbitmap %>%
  ggplot(aes(x=img_media_type, y=n_files, group=user_group, fill=user_group)) +
  geom_col(position = "dodge") +
  scale_color_brewer("User Group", palette = "Set1") +
  labs(x = "Media type", y = "No. of files", title = "File contributions on Wikimedia Commons by non-bitmap media types") +
  theme_minimal(base_size = 11) +
   theme(legend.position = "bottom", strip.placement = "outside", axis.text.x = element_text(angle = 90))

```


## File contributions by month 
```{r}
query <- "
SELECT LEFT(img_timestamp, 6) AS yr_month, user_group, COUNT(*) AS n_files
FROM (
-- Get active bots
SELECT ug_user AS user_id, ug_group AS user_group
FROM user_groups
WHERE ug_group = 'bot'
UNION
-- Get inactive bots
SELECT ufg_user AS user_id, ufg_group AS user_group
FROM user_former_groups
WHERE ufg_group = 'bot'
UNION
-- Get user ids with bot categories in their user pages
SELECT user.user_id, 'bot' AS user_group
FROM user INNER JOIN (
  -- all user page names with bot category
  SELECT REPLACE(page.page_title, '_', ' ') AS user_name
  FROM page INNER JOIN (
    -- page ids with bot categories 
    SELECT DISTINCT cl_from AS page_id
    FROM categorylinks
    WHERE cl_to REGEXP '_(bot_flag|bots)(_|$)'
      AND cl_type = 'page'
  ) AS bot_cat ON page.page_id=bot_cat.page_id
  WHERE page_namespace = 2
) AS bot_name ON user.user_name=bot_name.user_name
) AS bots RIGHT JOIN image ON bots.user_id = image.img_user
GROUP BY LEFT(img_timestamp, 6), user_group;
"
```

```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
bot_by_month <- wmf::mysql_read(query, "commonswiki", con = con)
message("Closing connection...")
wmf::mysql_close(con)
save(bot_by_month, file = "data/bot_by_month.RData")
```

```{r}
load("data/bot_by_month.RData")

bot_by_month <- bot_by_month %>%
  mutate(year_month = zoo::as.yearmon(yr_month, "%Y%m"),
         user_group = if_else(is.na(user_group), "user", user_group)) %>%
  arrange(year_month, user_group) %>%
  group_by(user_group) %>%
  mutate(total_files = cumsum(n_files)) %>%
  group_by(yr_month) %>%
  mutate(proportion = total_files / sum(total_files))
```

### Cumulative sum of file contributions by month 
```{r}
library("polloi")
bot_by_month %>%
  ggplot(aes(x=year_month, y=total_files, group=user_group, color=user_group)) +
  geom_line(size = 1) +
  zoo::scale_x_yearmon(breaks = seq(min(bot_by_month$year_month), max(bot_by_month$year_month), by = 1)) +
  scale_y_continuous(labels = polloi::compress) +
  scale_color_brewer("User Group", palette = "Set2") +
  labs(x = "Month", y = "Cumulative sum of files", title = "Cumulative sum of file contributions on Wikimedia Commons by users and bots") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", strip.placement = "outside", axis.text.x = element_text(angle = 90))

```



```{r}
bot_by_month %>%
  ggplot(aes(x=year_month, y=proportion, group=user_group, color=user_group)) +
  geom_line(size = 1) +
  zoo::scale_x_yearmon(breaks = seq(min(bot_by_month$year_month), max(bot_by_month$year_month), by = 1)) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_brewer("User Group", palette = "Set2") +
  labs(x = "Month", y = "Proportion of files (Cumulative)", title = "Proportion of files contributions on Wikimedia Commons by bots and users") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", strip.placement = "outside", axis.text.x = element_text(angle = 90))

```

### Newly uploaded (non-cumulative) file contributions by month
```{r}
 bot_by_month %>% 
        filter(yr_month != 201802) %>% 
#filter out last month since query was run at the beginning of February.
 ggplot(aes(x=year_month, y=n_files, group=user_group, color=user_group)) +
  geom_line(size = 1) +
  zoo::scale_x_yearmon(breaks = seq(min(bot_by_month$year_month), max(bot_by_month$year_month), by = 1)) +
  scale_y_continuous(labels = polloi::compress) +
  scale_color_brewer("User Group", palette = "Set2") +
  labs(x = "Month", y = "Number of new files", title = "Total number of newly uploaded files on Wikimedia Commons by users and bots") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", strip.placement = "outside", axis.text.x = element_text(angle = 90))
```

Note: The plot above show the total number of new file contributions to Commons each month but does not account for deleted files 


