---
title: "File deletion times on Wikimedia Commons (Redux)"
author: "MNeisler"
date: "2/9/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(0)
library(tidyverse)
```

```{r}
message("Create an auto-closing SSH tunnel in the background")
system("ssh -f -o ExitOnForwardFailure=yes stat1006.eqiad.wmnet -L 3307:analytics-store.eqiad.wmnet:3306 sleep 10")
library(RMySQL)
con <- dbConnect(MySQL(), host = "127.0.0.1", group = "client", dbname = "commonswiki", port = 3307)
```

Epic Ticket: [T185363](https://phabricator.wikimedia.org/T185363)

## File deletions on Wikimedia Commons - Breakdown by Time (Redux)

This report is a replicate of Mikhail's report in October 2017, reviewing the time to deletion for files uploaded on Wikimedia Commons. Questions reviewed included: 

1. Time to deletion, broken up by file type and reason for deletion (copyright violation vs. other).
2. The proportion of files deleted within a month after uploaded. How does this deletion rate look like over time? I looked at both proportion of all uploaded files (both deleted and not deleted before 2018-02-01) and just uploaded files that were deleted before 2018-02-01.

Ticket: [T18675](https://phabricator.wikimedia.org/T186575)


### 1. Time to deletion, broken up by file type and reason for deletion (copyright violation vs. other).

```{r}
query <- "SELECT 
        fa_media_type AS media_type,
(
    INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'copyvio') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'copyright') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'trademark') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'logo') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'fair use') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'dmca') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'non-free') > 0
    OR INSTR(LOWER(CONVERT(fa_deleted_reason USING utf8)), 'not free') > 0
  ) AS copyright_nonfree,
        fa_timestamp AS upload_date, 
        fa_deleted_timestamp AS delete_date
FROM filearchive"
```


```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
time_to_deletion <- wmf::mysql_read(query, "commonswiki", con = con)
message("Closing connection...")
wmf::mysql_close(con)
save(time_to_deletion, file = "data/time_to_deletion.RData")
```
```{r}
load("data/time_to_deletion.Rdata")
#Convert date to y-m-d format
library(magrittr)
library(lubridate)
time_to_deletion$reason <- if_else(time_to_deletion$copyright_nonfree == 1, "copyright, logo, or non-free", "other reason")
time_to_deletion$copyright_nonfree <- NULL
time_to_deletion$media_type %<>% tolower
time_to_deletion$upload_date %<>% ymd_hms(tz = "UTC")
time_to_deletion$delete_date %<>% ymd_hms(tz = "UTC")
time_to_deletion$difference <- as.numeric(difftime(time_to_deletion$delete_date, time_to_deletion$upload_date, units = "secs"))


time_to_deletion %<>%
        filter(difference > 0 & !is.na(media_type)) 

```

```{r}
#find median times
median_times <- time_to_deletion %>%
        group_by(reason) %>%
        summarise(median_time = ceiling(median(difference))) %>%
        spread(reason, median_time) %>%
        unlist
```

```{r}
logtime_breaks <- c(1, 60, 60*60, 60*60*24, 60*60*24*7, 60*60*24*28, 60*60*24*365, 60*60*24*365*10)
logtime_labels <- function(breaks) {
  lbls <- breaks %>%
    round %>%
    seconds_to_period() %>%
    tolower %>%
    gsub(" ", "", .) %>%
    sub("(.*[a-z])0s$", "\\1", .) %>%
    sub("(.*[a-z])0m$", "\\1", .) %>%
    sub("(.*[a-z])0h$", "\\1", .) %>%
    sub("(.*[a-z])0d$", "\\1", .)
  lbls <- case_when(
    lbls == "7d" ~ "1wk",
    lbls == "28d" ~ "1mo",
    lbls == "365d" ~ "1yr",
    lbls == "3650d" ~ "10yrs",
    TRUE ~ lbls
  )
  return(lbls)
}
scale_x_logtime <- function(...) {
  scale_x_log10(..., breaks = logtime_breaks, labels = logtime_labels)
}
scale_y_logtime <- function(...) {
  scale_y_log10(..., breaks = logtime_breaks, labels = logtime_labels)
}

```


```{r}
library(ggplot2)
p <- ggplot(
        filter(time_to_deletion, !media_type %in% c("unknown", "archive", "text")),
        aes(x = difference, fill = reason)
        ) +
        geom_density(adjust = 1.5, alpha = 0.5) +
        scale_x_logtime(name = "Time to deletion") +
        facet_wrap(~ media_type, scales = "free_y") +
        wmf::theme_facet(14) +
  theme(
          panel.grid.minor.x = element_blank(),
          axis.text.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          axis.title.y = element_blank()
          ) +
        labs(
                title = "Distribution of files' time to deletion by media type and reason for deletion",
                subtitle = paste("The median time to deletion across all media types is", median_times["copyright, logo, or non-free"], "seconds",
                                 "for copyright-related reasons and", median_times["other reason"], "otherwise"))

p 
ggsave("figures/time-to-deletion.png", p, width = 18, height = 9, units = "in", dpi = 150)
```


### 2. The proportion of files deleted within a month after uploaded. How does this deletion rate look like over time?

```{r}
#Find all files deleted within a month (28 days after upload date)
monthly_deletions <- time_to_deletion %>%
        mutate(upload_month = floor_date(upload_date, "month"),
               delete_month = floor_date(delete_date, "month")) %>% 
        filter(upload_month < "2018-02-01") %>%
        group_by(upload_month, media_type) %>% 
        mutate(delete_in_month = difference < 60*60*24*28) 

monthly_deletions$upload_month <- as.Date(monthly_deletions$upload_month)
#Revise delete in month to factor and clarfiy TRUE and FALSE labels. 
monthly_deletions$delete_in_month %<>% factor(c(TRUE, FALSE), c("Within 1 month of upload", "After 1 month since upload"))
```

```{r}
require(ggplot2)
require(scales)

p <- monthly_deletions %>%
  group_by(upload_month, delete_in_month) %>%
  tally %>%
        ggplot(aes(x=upload_month, y = n, fill = delete_in_month)) +
        geom_area(position = "fill", color = "black") +
        geom_hline(yintercept = 0.5, linetype = "dashed") +
        scale_y_continuous(labels = scales::percent_format()) +
        scale_x_date(date_labels = "%Y", date_breaks = "1 year", date_minor_breaks = "1 year")  +
        wmf::theme_min() +
        labs(
                fill = "Deleted",
                x= "Date",
                y = "Proportion",
                title = "Proportion of all uploaded files deleted within 1 month (28 days)",
                subtitle = "Includes only files deleted as of 2018-02-01"
  )
p
ggsave("figures/monthly_deletions_alldel.png", p, width = 18, height = 9, units = "in", dpi = 150)

```


```{r}
require(ggplot2)
require(scales)

p <- filter(monthly_deletions, !media_type %in% c("unknown", "archive", "text")) %>% 
        ggplot(aes(x=factor(1), fill=delete_in_month)) +
        geom_bar(position = "fill") +
        scale_y_continuous(labels=percent_format()) +
        facet_wrap(~ media_type, scale = "free_y") +
        labs(title = "Proportion of newly uploaded files deleted within 1 month by media type", 
             fill = "Deleted",
             x = "",
             y = "Proportion of files deleted within 1 month ",
        subtitle = "Includes only files deleted as of 2018-02-01"
p
ggsave("figures/monthly_deletions_bytype.png", p, width = 18, height = 9, units = "in", dpi = 150)
```


## Proportion of files deleted within a month of upload (Looking at all uploaded file each month [deleted and not deleted as of 2018-02-01])

The above plots only account for files deleted before 2018-02-01. I also reviewed the proportion of all uploaded files each month both deleted and not deleted as of 2018-02-01. 

```{r}
all_uploads_result <- "SELECT 
upload_date, COUNT(*) AS uploads
FROM(
SELECT
DATE(LEFT(filearchive.`fa_timestamp`, 8)) AS upload_date
FROM filearchive
UNION ALL
SELECT
DATE(LEFT(image.`img_timestamp`, 8)) AS upload_date
FROM image
) AS dates
GROUP BY upload_date;
"
```


```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
all_uploads <- wmf::mysql_read(all_uploads_result, "commonswiki", con = con)
message("Closing connection...")
wmf::mysql_close(con)
save(all_uploads, file = "data/all_uploads.RData")
```

```{r}
deleted_in_month_result <- "SELECT 
DATE(LEFT(fa_timestamp, 8)) AS upload_date, Count(*) AS n_deleted_month
FROM filearchive
WHERE DATE(TIME_TO_SEC(fa_timestamp)) - DATE(TIME_TO_SEC(fa_deleted_timestamp)) <= 2419200
GROUP BY DATE(LEFT(fa_timestamp, 8))
"
```


```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
deleted_in_month <- wmf::mysql_read(deleted_in_month_result, "commonswiki", con = con)
message("Closing connection...")
wmf::mysql_close(con)
save(deleted_in_month, file = "data/deleted_in_month.RData")
```

```{r}
load("data/all_uploads.Rdata")
load("data/deleted_in_month.Rdata")
monthly_deletions_all <- merge(all_uploads, deleted_in_month, by = "upload_date", all = TRUE)
```


```{r}
library(lubridate)
library(magrittr)
monthly_deletions_all$upload_date %<>% ymd()
#assign NAs to 0 as these are instances where no files uploaded were deleted within a month
monthly_deletions_all$n_deleted_month[is.na(monthly_deletions_all$n_deleted_month)] <- 0
tail(monthly_deletions_all)
```

```{r}
monthly_deletions_alluploads <- monthly_deletions_all %>%
        mutate(upload_month = floor_date(upload_date, "month")) %>%
        group_by(upload_month)  %>%
        summarise(uploads = sum(uploads),
                  n_deleted_month = sum(n_deleted_month)) %>%
        ungroup() %>%
        mutate(prop_deleted = n_deleted_month/uploads) %>%
        arrange(upload_month, uploads, n_deleted_month)

tail(monthly_deletions_alluploads)
```
```{r}
max_prop <- monthly_deletions_alluploads  %>% 
        top_n(4, prop_deleted)

max_prop

```

```{r}
p <- ggplot(filter(monthly_deletions_alluploads, upload_month < "2018-02-01"),
            aes(x = upload_month, y = prop_deleted)) +
        geom_line() +
        geom_point(data = max_prop) +
        geom_label(
                data = max_prop,
                aes(
                        label = paste0(round(prop_deleted, 4) * 100, "%"), hjust = "right", vjust = "top"
                ),
                size = 3
) + 
        geom_hline(aes(yintercept = median(prop_deleted)), linetype = "dashed") +
        scale_y_continuous(labels = scales::percent_format()) +
        scale_x_date(date_labels = "%Y", date_breaks = "1 year", date_minor_breaks = "1 year")  +
        wmf::theme_min() +
        labs(
                x= "Date of upload",
                y = "Proportion deleted within 1 month",
                title = "Proportion of all uploaded files deleted within 1 month (28 days)",
                subtitle = "Includes all uploaded files both deleted and not deleted as of 2018-02-01"
  )
p
ggsave("figures/monthly_deletions_alluploads.png", p, width = 18, height = 9, units = "in", dpi = 150)
```
```{r}
(mean(monthly_deletions_alluploads$prop_deleted))*100
```

