---
title: "Distribution of file type and trends on Wikimedia Commons (Redux)""
author: "MNeisler"
date: "2/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(0)
library(tidyverse)
```


```{r}
message("Create an auto-closing SSH tunnel in the background")
system("ssh -f -o ExitOnForwardFailure=yes stat1006.eqiad.wmnet -L 3307:analytics-store.eqiad.wmnet:3306 sleep 10")
library(RMySQL)
con <- dbConnect(MySQL(), host = "127.0.0.1", group = "client", dbname = "commonswiki", port = 3307)
```

Epic Ticket: [T185363](https://phabricator.wikimedia.org/T185363)

## Distribution of file types and trends on Wikimedia Commons

This report is a replicate of Mikhail's report in October 2017, reviewing the distribution of file types and extensions. I also reviewed the the cumulative upload counts and newly uploads per month by file extension. 

Questions addressed:
1. The distribution of file types and extensions. 

2. Cumulative upload counts and newly uploads per month by file extension.

For the purpose of this analysis, I defined "newly uploaded" as files that were added and not deleted from upload until today. 

Ticket: [T18675](https://phabricator.wikimedia.org/T186575)


### File type table
I first queried the image table of the commons wiki to select the img_timestamp and file extension type (jpeg, gif, png, etc). The file name was used to determine the format of the file as the "img_minor_mime" and "img_media_type" fields may not be accurately labeled for all files. 

```{r}
query <- "SELECT
  DATE(LEFT(img_timestamp, 8)) AS `date`,
  SUBSTRING(REGEXP_SUBSTR(LOWER(CONVERT(img_name USING utf8)), '\\\\.([a-z]{3,})$'), 2, 10) AS extension,
  COUNT(*) AS uploads
FROM image
GROUP BY `date`, extension;"
```

```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
uploads_by_filetype <- wmf::mysql_read(query, "commonswiki", con)
message("Closing connection...")
wmf::mysql_close(con)
save(uploads_by_filetype, file = "data/uploads_by_filetype.RData")
```

```{r}
load("data/uploads_by_filetype.Rdata")
#Convert date to y-m-d format
library(magrittr)
library(lubridate)
uploads_by_filetype$date %<>% ymd()
#remove empty extension fields.
uploads_by_filetype <- uploads_by_filetype[uploads_by_filetype$extension != "", ]
```


```{r}
#rename extension files to group jpg/jpeg and tif/tiff files together.
uploads_by_filetype %<>%
  mutate(extension = case_when(
    extension %in% c("jpg", "jpeg") ~ "jpg/jpeg",
    extension %in% c("tif", "tiff") ~ "tif/tiff",
    TRUE ~ extension
  ))
```

### Find the cumulative file uploads by month and file type
```{r}
#Sum of uploads by file extension type. 
total_uploads <- uploads_by_filetype %>%
        group_by(extension) %>%
        summarize(uploads = sum(uploads)) %>%
        arrange(desc(uploads))
```

```{r}
# Find cumulative uploads column broken down by extension and month. 
cumulative_uploads <- 
        uploads_by_filetype %>%
        group_by(extension) %>%
        arrange(date) %>%
        mutate(cumulative_uploads = cumsum(uploads)) %>%
        ungroup() %>%
        mutate(extension = factor(extension,
                                  total_uploads$extension))
```


```{r}
library(ggplot2)
p <- ggplot(filter(cumulative_uploads, date < "2018-02-01"), aes(x = date, y = cumulative_uploads)) +
  geom_line() +
  scale_y_continuous("Cumulative files uploaded", labels = polloi::compress) +
  scale_x_date("Date", date_breaks = "2 years", date_minor_breaks = "1 year", date_labels = "'%y") +
  facet_wrap(~ extension, scales = "free_y") +
  wmf::theme_facet(14) +
  labs(
    title = "Wikimedia Commons cumulative upload counts by file extension type",
    subtitle = "Does not include files that have been deleted as of 2018-02-11"
  )

p 
ggsave("figures/cumulative_uploads.png", p, width = 18, height = 9, units = "in", dpi = 150)
```

### Find the monthly file uploads by month and file type
```{r}
monthly_uploads <- uploads_by_filetype %>%
        mutate(month = floor_date(date, "months")) %>% #round down to months
        group_by(month, extension) %>%
        summarize(monthly_uploads = sum(uploads)) %>%
        ungroup()

head(monthly_uploads)
```
```{r}
#Find the top monthly uploads for each file type. 
top_upload_months <- monthly_uploads %>%
        filter(month < "2018-02-01") %>%
        group_by(extension) %>%
        top_n(1, monthly_uploads)

```


```{r}
p <- ggplot(filter(monthly_uploads, month < "2018-02-01"), 
            aes(x = month, y = monthly_uploads)) +
        geom_line() +
        geom_point(data = top_upload_months) +
        geom_label(
                data = top_upload_months,
                aes(
                        label = paste(polloi::compress(monthly_uploads, 0),
                                      "in", format(month, "%b %Y")),
                        hjust = "right", vjust = "top"
                        ),
                size = 3) +
        scale_y_continuous("Uploads per month", labels = polloi::compress) +
        scale_x_date("Date", date_breaks = "2 years", date_minor_breaks
                     = "1 year", date_labels = "'%y") +
        facet_wrap(~ extension, scales = "free_y") +
        wmf::theme_facet(14) +
        labs(
                title = "Wikimedia Commons monthly file upload counts by file extension",
                subtitle = "Does not include files that have been deleted as of 2018-02-11")

p
ggsave("figures/monthly_uploads_byfiletype.png", p, width = 18, height = 9, units = "in", dpi = 150)

```

### Proportions of file distribution on Wikimedia Commons by media and file type

```{r}
extension_by_media <- bind_rows(list(
  "image" = data_frame(extension = c("jpg/jpeg", "png", "svg", "tif/tiff", "gif", "xcf", "webp")),
  "audio" = data_frame(extension = c("flac", "mid", "wav", "ogg", "oga", "opus")),
  "video" = data_frame(extension = c("ogv", "webm")),
  "document" = data_frame(extension = c("pdf", "djvu"))
), .id = "media")

total_uploads %<>%
        left_join(extension_by_media, by = "extension")
```

```{r}
library(viridis)
library(treemapify)

p <- ggplot(
        filter(total_uploads, extension != "jpg/jpeg"),
        aes(area = uploads, fill = log10(uploads + 1), label =
                    extension, subgroup = media)
        ) +
        geom_treemap(color = "black") +
        geom_treemap_subgroup_border(color = "black") +
        geom_treemap_text(color = "black", place = "topleft", grow =
                                  FALSE, min.size = 0) +
        geom_treemap_subgroup_text(
                place = "center", grow = TRUE, alpha = 0.5, color =
                        "black",
                fontface = "italic", min.size = 0
                ) +
        scale_fill_viridis("Total files uploaded (on a logarithmic scale)", labels = function(x) {return(polloi::compress(10 ^ x))
                }) +
        guides(fill = guide_colorbar(barwidth = 20, barheight = 1))+ wmf::theme_min(14) +
        labs(
    title = "Proportion of cumulative files uploaded to Wikimedia Common by file and media type",
    caption = "Note: Jpg/jpeg file extensions were omitted for visual clarity as they represent over ~38M of the files uploaded.",
    subtitle = "These 16 extensions make up 5.7M files that have been uploaded and not deleted as of 2018-02-11"
  )
p
ggsave("figures/treemap_uploads.png", p, width = 9, height = 9, units = "in", dpi = 150)
```


### Summary of total uploads by file type

```{r}
library(knitr)
total_uploads %>%
  select(media, extension, uploads) %>%
  arrange(desc(uploads)) %>%
  kable(format = "markdown")
```


