---
title: "File deletions by users on Wikimedia Commons (Redux)"
author: "MNeisler"
date: "2/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(0)
library(tidyverse)
```

```{r}
message("Create an auto-closing SSH tunnel in the background")
system("ssh -f -o ExitOnForwardFailure=yes stat1006.eqiad.wmnet -L 3307:analytics-store.eqiad.wmnet:3306 sleep 10")
library(RMySQL)
con <- dbConnect(MySQL(), host = "127.0.0.1", group = "client", dbname = "commonswiki", port = 3307)
```

Epic Ticket: [T185363](https://phabricator.wikimedia.org/T185363)

## File deletions by user on Wikimedia Commons (Redux)

This report is a replicate of Mikhail's report in October 2017, reviewing file deletions on Wikimedia Commons. Questions reviewed in this report include: 

1. Number of deleters (users who have deleted at least one file) over time
2. How many files each user has deleted?

Ticket: [T18675](https://phabricator.wikimedia.org/T186575)


###  Number of deleters (users who have deleted at least one file) over time

```{r}
query <- "SELECT
  fa_deleted_timestamp,
  fa_deleted_user,
  user_groups.ug_group
FROM filearchive
LEFT JOIN user_groups ON filearchive.fa_deleted_user = user_groups.ug_user;"
```

```{r}
message("Using SSH tunnel & connection to Analytics-Store...")
deleters <- wmf::mysql_read(query, "commonswiki", con)
message("Closing connection...")
wmf::mysql_close(con)
save(deleters, file = "data/deleters.RData")
```

```{r}
load("data/deleters.Rdata")
library(magrittr)

deleters %<>%
  rename(
    delete_date = fa_deleted_timestamp,
    user_id = fa_deleted_user,
    group = ug_group
  ) %>%
   mutate(
    delete_date = ymd_hms(delete_date),
    # Anonymize users
    user = as.numeric(factor(as.character(user_id)))
  ) %>%
  select(-user_id) #Drop user ID
```

```{r}
deletion_counts <- deleters %>%
   mutate(date = as.Date(delete_date)) %>%
   group_by(user, date) %>%
   summarize(
    deletions = n(),
    groups = paste0(group, collapse = ", ")
  )
```


```{r}
#Count the number of users who've deleted at least one file.
user_counts_cumulative <- deletion_counts %>%
        group_by(user) %>% 
        #first deletion date
        summarize(first_deletion = min(date)) %>% 
        arrange(first_deletion) %>% 
        group_by(first_deletion) %>% 
        summarise(new_users = n()) %>% 
        mutate(cumulative_deleters = cumsum(new_users))

tail(user_counts_cumulative)
```


### Cumulative Sum of Deleters on Wikimedia Commons
```{r}
library(ggplot2)
p <- ggplot(user_counts_cumulative) +
  geom_line(aes(x = first_deletion, y = cumulative_deleters)) +
  scale_x_date("Date", date_breaks = "1 year", date_labels = "'%y") +
  scale_y_continuous("Users", breaks = seq(0, 600, 100)) +
  wmf::theme_min(14, panel.grid.minor.x = element_blank()) +
  labs(
    title = "Total No. of Deleters on Wikimedia Commons",
    subtitle = "Number of users who have deleted at least one file"
  )
ggsave("figures/cumulative_deleters.png", p, width = 6, height = 3, units = "in", dpi = 150)
```

### Deleters broken down by quantity intervals

```{r}
library(knitr)
user_counts_byquantity <- deletion_counts %>%
        group_by(user) %>%
        summarize(deletions = sum(deletions))  %>%
        #Divide range of deletions into intervals and convert to character
        mutate(deletions = as.character(cut(
                deletions,
                breaks = c(0, 10, 50, 100, 500, 1e3, 5e3, 1e4, Inf))
                )) %>%
        group_by(deletions) %>%
        #count the users in each interval
        summarize(users = n()) %>%
        mutate(deletions = factor(case_when(
                deletions == "(0,10]" ~ "0-10",
                deletions == "(10,50]" ~ "10-50",
                deletions == "(50,100]" ~ "50-100",
                deletions == "(100,500]" ~ "100-500",
                deletions == "(500,1e+03]" ~ "500-1K",
                deletions == "(1e+03,5e+03]" ~ "1K-5K",
                deletions == "(5e+03,1e+04]" ~ "5K-10K",
                deletions == "(1e+04,Inf]" ~ "10K+"), 
                c("0-10", "10-50", "50-100", "100-500", "500-1K", "1K-5K", "5K-10K", "10K+"))) 
        
```

```{r}
p <- ggplot(user_counts_byquantity, aes(x = deletions, y = users)) +
        geom_bar(stat = "identity") +
        geom_text(aes(
                label = sprintf("%.0f of users", users),
                vjust = "bottom"
                ), nudge_y = 5) +
        wmf::theme_min(14) +
        labs(
                title = "File deletion activity on Wikimedia Commons",
                subtitle = sprintf("%.0f users have collectively deleted %s files",
                                   max(deletion_counts$user),
                                   polloi::compress(sum(deletion_counts$deletions), 2)
                                   ),
                x = "Number of files each user has deleted",
                y = "Users who have deleted this many files")
ggsave("figures/deleter_activity.png", p, width = 12, height = 6, units = "in", dpi = 150)
```

